Characterization of gaps in directly bonded Si compound optics

### 2014-03-31 AO_bondedSi.tex
Add corrections from DTJ up to page 4

\begin{abstract}
Si direct bonding offers flexibility in the design and development of large Si optics.  The bonding process presents challenges in meeting the performance demands and physical robustness, in particular for cryogenic and space applications.  Air cavities in between the Si interface will have large Fresnel reflections which will interfere as a low-Finesse Fabry-P\`erot etalon.  Even small 35 nm gaps reduce transmission through a direct bonded Si compound optic by 5\% at $\lambda = $ 1.15 $\mu$m at normal incidence.  Stock near infrared spectrophotometers like the Cary5000 from Agilent offer 0.2\% photometric precision.  We can detect measure gaps as small as 8 nm by measuring transmission as a function of wavelength.  Our approach involves modeling multiple incoherent reflections within the substrate interfaces with a wave transfer matrix, modified for intensities rather than complex amplitudes.  The detection and morphology of minuscule gaps provides valuable feedback in the design of Si direct bonding strategies.
\end{abstract}

@@

Direct optical bonding of silicon allows manufacturers to combine parts with no Fresnel loss at the interface.  The high refractive index of Si makes this advantage particularly useful.  Traditional figures of merit for the optical audits of bonds, however, are either insensitive to gaps that are small yet have a significant impact on throughput and phase coherence, or do not provide quantitative information about the extent of the gaps along the direction of propagation.

Issues with the quality of the bonds become particular acute when the optics are used in double-pass and/or when the incidence angle at the bonds is far from the normal.

The University of Texas diffractive optics group produces diffraction gratings on silicon using microlithographic techniques \cite{1998SPIE.3354..201J,2010SPIE.7739E.146W}.  These gratings are used either in transmission (i.e. as grisms \cite{2008SPIE.7014E..77D, 2010SPIE.7739E.123G}), or in reflection (i.e. as immersion gratings, \cite{2007ApOpt..46.3400M, 2010SPIE.7739E.146W,2012SPIE.8450E..2SG}.  These gratings operate throughout the infrared, from 1.2 to 38 $\mu$m, and find application in ground-based (IGRINS \cite{2010SPIE.7735E..54Y, 2012SPIE.8450E..2SG}), airborne (SOFIA FORCAST \cite{2008SPIE.7014E..77D}), and space (JWST NIRCAM \cite{2005SPIE.5904...21B,2010SPIE.7739E.123G}) applications.  To date, our group has produced the gratings on large (up to 100 mm diameter and 30 mm thick) monolithic pieces of monocrystalline silicon and we shape the substrates by sawing, grinding, and polishing to form the final devices \cite{2010SPIE.7739E.146W}.  The next generation of astronomy and earth science spectrographs, however, will need optical devices larger and more precise than what can be made with existing technology.  The proposed Giant Magellan Telescope \cite{2012SPIE.8444E..1HJ} Near Infrared Spectrograph \cite{2006SPIE.6269E.143J,2010SPIE.7735E..87L} (GMTNIRS) will need immersion gratings with thicknesses up to 60 mm and lengths up to 200 mm.  Substrates of the required size cannot fit into conventional processing equipment like reactive ion etchers and electron-beam lithography tools.  Large, thick Si substrates are also prohibitively expensive to experiment on.  Silicon direct bonding \cite{1986JAP....60.2987S,2012SPIE.8450E..2TV} offers advantages by reducing the thickness and mass needed for the development of a thin (0.5$-$10 mm) Silicon diffractive optic that can later be bonded to a fine polished Si prism substrate.  Other groups producing SI diffractive optics are also experimenting with this approach (see \cite{2012SPIE.8450E..2TV} and \url{www.lumella.com}).

While it is possible that proper precautions to eliminate all micron-sized particles from both SI surfaces prior to bonding, the power-law nature of the particle size distribution in clean-room environments (\cite{Cooper_1986}) means that it is significantly more difficult to assure that no particles of the 10$-$60 nm size scale are present.  Small particles could be a concern if they act as "tent poles", ti create gaps where the bonding is incomplete.

There are two major concerns with Si direct bonding: the performance and robustness.  Specifically, what happens if the bonding process is imperfect and there are small gaps between the Si parts?  Does a gap degrade the performance of a final device, either through optical aberrations or transmission loss?  How mechanically robust is the final unit?  What is the strength of the bond?  Can the interface be broken by shearing, torquing, pulling, or thermal stress?  In practice, the figure of merit for robustness is whether the part can survive a shake test simulating the forcing experienced during a rocket launch or transport, or during cryogenic cycling, if the part will be used at low temperature.  Optically, the relevant parameters are the loss caused by the gap and the effective distortion of the transmitted wavefront.

There are two main solutions for dealing with bubbles.  We will refer to the two approaches as the \emph{strength approach} and the \emph{resilience approach}.  The \emph{strength approach} is to prevent bubbles from forming in the first place, specifically by removing all possible contaminants from the environment before bonding.  Typically this approach demands disciplined handling protocols, and rigorous multi-stage cleaning steps, although unconventional micro-cleanroom approaches exist \cite{1989JaJAP..28L2141L}.  The \emph{resilience approach} is simply to deal with the uncomfortable likelihood that bubbles will occur, and then do something to remove them.  The most cited way to get rid of IR bubbles is to anneal them\cite{Mitani1990}.  We describe the current understanding of both approaches in section \ref{secHistory}.

This paper introduces a technique for directly measuring gaps at silicon-silicon interfaces.  Other techniques for measuring and mapping gaps have already been described elsewhere \cite{1992JEMat..21..669M}, including IR interference\cite{Mitani1990}, X-ray topography\cite{1994JaJAP..33....6H}, and ultrasound microscopy\cite{2000RScI...71.1869G}.  The standard infrared interference technique is to examine the bond in reflection for the presence of Newton rings.  The presence and number of such rings can yield information about both the lateral and axial extend of any gap, as long as the gap along the direction of propagation is of the order of the vacuum wavelength or larger.  X-ray topography \cite{1992JEMat..21..669M,1994JaJAP..33....6H} exploits the change in optical density along lines of sight with gaps to allow precise and high spatial resolution measurements of the spatial extent of gaps with axial dimensions as small as a few nanometers but provides only coarse estimates of the axial extent.  Ultrasound microscopy exploits the density change at the gap interface to produce reflections of sound waves and map the lateral gap dimensions but produces limited information about the axial dimension \cite{2000RScI...71.1869G}.

Here we describe an exploitation of the IR interference technique with new precision instrumentation in the near infrared wavelength range to measure the axial extent of gaps at the few nanometer size scale.  In this paper we show that gaps as small as 35 nm cause an unacceptable degradation in transmission performance.  Because the size scales probed are so small, they are comparable to the range over which molecular (van der Waals) forces are acting.  It is conceivable that the strength of the bond energy can be constrained from the \emph{in situ} measurement of the separation and that therefore the optical technique can provide not only a measure of the transmission but also an indication of the mechanical robustness of the bond.  By measuring both the spatial and axial extent of any gaps, the new metrology technique informs revisions of the bonding strategy that can improve both the optical and mechanical performance.

Our new technique for measurement of the axial extent of gaps exploits wavelength-dependence of reflection losses at the gap and the availability of commercial scanning monochromator systems with both high precision and good amplitude stability.  We first describe the theory behind the technique and predicted outcomes for various gap dimensions.  We then evaluate the technique experimentally on a variety of bonded samples.

@@

The foundation paper on Si bonding came in 1986\cite{1986JAP....60.2987S}.  After that, there was a sequence of papers using IR imaging to look for evidence of defects in the bond.  In some of cases, the bond propagation could be directly monitored\cite{1988JaJAP..27L2364S, 1995ApPhL..67.3614G}.  The bond process could be shown to be bubble-free\cite{1988JaJAP..27L2364S, 1989JaJAP..28L2141L}.  The bond surface energy was directly measured\cite{1989JaJAP..28L2141L}.  The causes and remedies for bubble formation were discussed as early as 1990\cite{Mitani1990}.  Specifically, the IR bubble density changes as a function of anneal temprature\cite{1992JEMat..21..669M}, with bubbles disappearing at temperatures greater than 1100$^\circ$ C.  The chemistry of the bonding interface began to be understood with infrared absorption spectroscopy and multiple internal reflection spectroscopy\cite{feijoo1994}, which lent credibility to the idea of distinct temperature phases and evidence for Si$-$H bonds\cite{1995ApPhA..61..101R}.  A robust picture of the chemistry as a function of anneal temperature was ultimate fleshed out\cite{1996JaJAP..35.2102R, 1998AnRMS..28..215G}.  Other variables potentially affecting bond strength were studied, for example surface roughness\cite{JJAP.37.4197}, surface topography \cite{2001JOptA...3...85G}, surface preparation\cite{1996ApPhL..68.2222T}, annealing time \cite{2000JAP....88.4404H}, ambient pressure and substrate thickness \cite{1995ApPhL..67..863G, 2007ApOpt..46.6793H}, and moisture and defects \cite{2001JAP....89.6013L}.  No where in the literature did we find a technique similar to the one presented here, probably because the technique relies on the high precision of relatively new IR capable spectroscopy that wasn't widely available until this decade.

>>>>

\begin{abstract}
Si direct bonding offers flexibility in the design and development of large Si optics.  The bonding process presents challenges in meeting the performance demands and physical robustness, in particular for cryogenic and space applications.  Air cavities in between the Si interface will have large Fresnel reflections which will interfere as a low-Finesse Fabry-P\`erot etalon.  Even small (35 nm) gaps reduce transmission through a direct bonded Si compound optic by 5\% at $\lambda = $ 1.15 $\mu$m at normal incidence.  Stock near infrared spectrophotometers like the Cary5000 from Agilent offer 0.2\% photometric precision.  We can use instruments like the Cary 5000 to detect and measure gaps as small as 8 nm by measuring transmission as a function of wavelength.  Our approach involves modeling multiple incoherent reflections within the substrate interfaces with a wave transfer matrix, modified for intensities rather than complex amplitudes.  The detection and morphology of minuscule gaps provides valuable feedback in the design of Si direct bonding strategies.
\end{abstract}

@@

Direct optical bonding of silicon allows manufacturers to combine Si parts with no Fresnel loss at the interface.  The high refractive index of Si makes this advantage particularly useful.  Traditional figures of merit for the optical audits of bonds, however, are either insensitive to gaps that are small yet have a significant impact on throughput and phase coherence, or do not provide quantitative information about the extent of the gaps along the direction of propagation.

%Issues with the quality of the bonds become particularly acute when the optics are used in double-pass and/or when the incidence angle at the bonds is far from the normal.

The University of Texas diffractive optics group produces diffraction gratings on silicon using microlithographic techniques \cite{1998SPIE.3354..201J,2010SPIE.7739E.146W}.  These gratings are used either in transmission (i.e. as grisms \cite{2008SPIE.7014E..77D, 2010SPIE.7739E.123G}), or in reflection (i.e. as immersion gratings, \cite{2007ApOpt..46.3400M, 2010SPIE.7739E.146W,2012SPIE.8450E..2SG}.  These gratings operate throughout the infrared, from 1.2 to 38 $\mu$m, and find application in ground-based (IGRINS \cite{2010SPIE.7735E..54Y, 2012SPIE.8450E..2SG}), airborne (SOFIA FORCAST \cite{2008SPIE.7014E..77D}), and space (JWST NIRCAM \cite{2005SPIE.5904...21B,2010SPIE.7739E.123G}) applications.  To date, our group has produced the gratings on large (up to 100 mm diameter and 30 mm thick) monolithic pieces of monocrystalline silicon and we shape the substrates by sawing, grinding, and polishing to form the final devices \cite{2010SPIE.7739E.146W}.  The next generation of astronomy and earth science spectrographs, however, will need optical devices larger and more precise than what can be made with existing technology.  The proposed Giant Magellan Telescope \cite{2012SPIE.8444E..1HJ} Near Infrared Spectrograph \cite{2006SPIE.6269E.143J,2010SPIE.7735E..87L} (GMTNIRS) will need immersion gratings with thicknesses up to 60 mm and lengths up to 200 mm.  Substrates of the required size cannot fit into conventional processing equipment like reactive ion etchers and electron-beam lithography tools.  Large, thick Si substrates are also prohibitively expensive to experiment on.  Silicon direct bonding \cite{1986JAP....60.2987S,2012SPIE.8450E..2TV} offers advantages by allowing us to produce the diffraction grating on a thin (0.5$-$10 mm) low-mass Silicon puck that can later be bonded to a fine polished Si prism substrate.  Other groups producing Si diffractive optics are also experimenting with this approach (see \cite{2012SPIE.8450E..2TV} and \url{www.lumella.com}).

While it is possible that proper precautions to eliminate all micron-sized particles from both Si surfaces prior to bonding, the power-law nature of the particle size distribution in clean-room environments \cite{doi:10.1080/02786828608959094} means that it is significantly more difficult to assure that no particles of the 10$-$60 nm size scale are present.  Small particles could be a concern if they act as "tent poles", that create gaps where the bonding is incomplete (Figure \ref{figParticle}).  Authors have referred to these large ($\sim$ 1 $\mu$m size) gaps as \emph{IR bubbles}, or simply bubbles, owing to their appearance in infrared (IR) transmission images as contiguous regions of a deficit of IR transmission.  They have also been referred to as voids, or interfacial voids, or gaps.  In this paper we refer to them as gaps.

Finite gaps will degrade the performance of a final device, either through optical aberrations or transmission loss.  The effect of the gap size on the IR transmission is the topic of this paper.  The effect of the gap on optical aberrations is beyond the scope of this paper.  A gap might also degrade the mechanical integrity of a final compound optic.  In practice, the part must survive thermal shocks and shake tests.  The mechanical properties of the gap are also beyond the scope of this paper.

There are two main solutions for reducing risk attributable to gaps.  We will refer to the two approaches as the \emph{strength approach} and the \emph{resilience approach}.  The \emph{strength approach} is to prevent bubbles from forming in the first place, specifically by removing all possible contaminants from the environment before bonding.  Typically this approach involves disciplined handling protocols, and rigorous multi-stage cleaning steps.  The reason the strength approach is so hard is that hydrocarbons serve as catalysts for gap formation\cite{1991JaJAP..30..615M}.  Hydrocarbons are common in Si wafer handling equipment, so avoiding hydrocarbons is challenging.  Unconventional micro-cleanroom approaches have been effective at meeting this goal \cite{1989JaJAP..28L2141L}.  The \emph{resilience approach} is simply to accept the uncomfortable likelihood that bubbles will occur, and then do something to remove them.  The most cited way to get rid of gaps is to anneal them\cite{Mitani1990}.  We describe the current understanding of both approaches in Section \ref{secHistory}.

Whatever technique one uses to remove or reuce the effects of bubbles, it is essential to have a quantitative method by which to assess accurately the optical impact of the gaps before and after treatment.  This paper introduces a technique for directly measuring gaps at silicon-silicon interfaces.  Other techniques for measuring and mapping gaps have already been described elsewhere \cite{1992JEMat..21..669M}.  One key technique is IR Newton rings from IR imaging\cite{Mitani1990}.  This paper gives deeper understanding of the infrared interference technique seen in IR images as Newton rings.  The number and morphology of Newton rings yields information about the lateral and axial extent of gaps. Specifically, dark-light fringes manifest when the gap is comparable to the vacuum wavelength or larger.  Our technique is IR interference in the limit of small gaps.

There are other techniques for mapping gaps.  X-ray topography \cite{1992JEMat..21..669M,1994JaJAP..33....6H} exploits the change in optical density along lines of sight with gaps to allow precise and high spatial resolution measurements of the spatial extent of gaps with axial dimensions as small as a few nanometers but provides only coarse estimates of the axial extent.  Ultrasound microscopy exploits the density change at the gap interface to produce reflections of sound waves and map the lateral gap dimensions but produces limited information about the axial dimension \cite{2000RScI...71.1869G}.

Here we describe an exploitation of the IR interference technique with new precision instrumentation in the near infrared to measure the axial extent of gaps at the few nanometer size scale.  In this paper we show that gaps as small as 35 nm cause an unacceptable degradation in transmission performance.  Because the size scales probed are so small, they are comparable to the range over which molecular (van der Waals) forces are acting.  It is conceivable that the strength of the bond energy can be constrained from the \emph{in situ} measurement of the separation and that therefore the optical technique can provide not only a measure of the transmission but also an indication of the mechanical robustness of the bond.  We do not pursue further the question of mechanical robustness.  By measuring both the spatial and axial extent of any gaps, the new metrology technique informs revisions of the bonding strategy that can improve both the optical and mechanical performance.

@@

The foundation paper on Si bonding came in 1986\cite{1986JAP....60.2987S}.  After that, there was a sequence of papers using IR imaging to look for evidence of defects in the bond.  \cite{1988JaJAP..27L2364S, 1995ApPhL..67.3614G} directly monitored the bond propagation with IR imaging.  \cite{1988JaJAP..27L2364S, 1989JaJAP..28L2141L} demonstrated gap-free bonding.  \cite{1989JaJAP..28L2141L} directly measured the bond surface energy.  \cite{Mitani1990} discussed the causes and remedies for bubble formation.  Specifically, the IR bubble density changes as a function of anneal temperature\cite{1992JEMat..21..669M}, with bubbles disappearing at temperatures greater than 1100$^\circ$ C.  The chemistry of the bonding interface began to be understood with infrared absorption spectroscopy and multiple internal reflection spectroscopy\cite{feijoo1994}, which lent credibility to the idea of distinct temperature phases and evidence for Si$-$H bonds\cite{1995ApPhA..61..101R}.  A robust picture of the chemistry as a function of anneal temperature was ultimately fleshed out\cite{1996JaJAP..35.2102R, 1998AnRMS..28..215G}.  Other variables potentially affecting bond strength were studied, for example surface roughness\cite{JJAP.37.4197}, surface topography \cite{2001JOptA...3...85G}, surface preparation\cite{1996ApPhL..68.2222T}, annealing time \cite{2000JAP....88.4404H}, ambient pressure and substrate thickness \cite{1995ApPhL..67..863G, 2007ApOpt..46.6793H}, and moisture and defects \cite{2001JAP....89.6013L}.  Notably, \cite{JJAP.37.4197} found that surface roughness above 1.3 nm RMS results in poor bonding quality in their sample of wafers etched with Argon.  Results from studies of surface topography are inconclusive\cite{2001JOptA...3...85G}.  Nowhere in the literature did we find a technique similar to the one presented here, probably because the technique relies on the high precision of relatively new IR capable spectroscopy that wasn't widely available until this decade.
